### 2024-01-02 thesis/MSECE_thesis.tex
Structural edits and proofreading edits.

\begin{document}
% Riddle thesis title page
\begin{center}
    \LARGE
    \textbf{
        \textit{
            Language Modeling of Aviation Communications\\
        }
    }

@@

\section{Abstract}\label{sec:abstract}
With recent advances in machine learning and deep learning technologies and the creation of larger aviation-specific corpora, applying
natural language processing technologies to aviation communications, especially those based on transformer neural networks, is becoming
increasingly feasible. Previous work has focused on machine learning applications to natural language processing, such as N-grams and
word lattices. This thesis will lay out a process for pretraining transformer-based language models on aviation English corpora and
compare the effectiveness and performance of language models transfer learned from pretrained checkpoints and those trained from their
base weight initializations (trained from scratch). The results show that transformer language models trained from scratch outperform
models fine-tuned from pretrained checkpoints. The work concludes by recommending future work to improve pretraining performance and
suggestions for downstream, in-domain tasks such as semantic extraction, named entity recognition (callsign identification), speaker
role identification, and speech recognition.

\section{Introduction}\label{sec:introduction}
There have been noticeable and significant advancements in the general domain of artificial intelligence (AI) recently. Natural
language processing and language modeling technologies, in particular, have made tremendous advances over the past two decades
\cite{hannun_deep_2014,lee_unsupervised_2009,li_jasper_2019,kriman_quartznet_2020}, achieving performance levels good enough to be brought to
consumer-facing products such as personal assistants, search engines, phones, televisions, etc. Due to the success in the general domain, applications
have begun to spin off into problem-specific domains, such as (and most notably for this work) aviation. For example, several speech recognition
systems with rule- and machine learning-based natural language processing, inference, and semantic extraction algorithms have been implemented at
EUROCONTROL simulation centers in Europe to train air traffic controllers and study the potential effects of AI-augmented air traffic management (ATM)
systems on controller workloads. Preliminary studies have found that AI-augmented ATM systems reduce controller workloads and fuel consumption of
aircraft by optimizing aircraft routes in the airspace and reducing departure/arrival times, thereby reducing fuel consumption
\cite{helmke_quantifying_2017}.

At the time of this writing, the current focus of general machine learning and deep learning algorithms in the language modeling domains has been on a
variety of deep learning methods/architectures, namely, recurrent, convolutional, and transformer neural networks, due to the availability of large
natural language corpora such as WikiText \cite{merity_pointer_2016} and IMDB \cite{maas_learning_2011}. In contrast to the general domain, the
aviation domain still prefers and sees much success with traditional machine learning models for language modeling applications. Typically, the most
prevalent of the traditional models are N-grams, word lattices, and rule-based approaches, depending on the application and operating environment of
the models. The most likely reason for this divergence between the domains is the difference and relative lack of labeled data in the aviation domain.
By comparison, the largest corpora in the general domain are either WikiText or the BookCorpus, which contain trillions of tokens, whereas, for
aviation, the largest corpus is Air Traffic Control Complete with approximately 26 hours of labeled data amounting to just over 300,000
tokens\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are labeled.}. This disparity
in data availability is exacerbated by the fact that aviation English uses a specialized vocabulary in addition to specialized pronunciation for some
words to enhance understandability in a radiotelephonic (R/T) medium \cite{paltridge_handbook_2013}. This makes the data labeling process
significantly more difficult because it not only requires two types of data (audio and text) to create labels, but also requires labelers to have
domain knowledge and experience to create effective transcriptions. This, in addition to the following facets, further contributes to the difficulty
of transcription: (1) the medium of communication, (2) the environment in which the communication occurs\footnote{For example, a noisy and possibly
    low fidelity medium, depending on the equipment used to transmit and receive communications}, (3) a noisy and active communication environment
i.e.~an active flight deck under the influence of wind, engine noise, etc., (4) an active Air Traffic Control (ATC) tower and channels, and (5) a
rapid rate of speech. Pilots are also operating sophisticated equipment during while communicating, requiring significant attention and mental
resources further contributing to the complexity of the communications \cite{paltridge_handbook_2013}.

@@

The key component missing for both of these applications is an effectively pretrained language model that can be used for these downstream tasks, so
the goal of this thesis is to run several experiments using transformer-based language models to determine the most effective strategy for pretraining
the model for downstream tasks. Recommended best practice is to use a pre-existing language model checkpoint and fine-tune on additional data
\cite{devlin_bert_2019,liu_roberta_2019,lewis_bart_2019}, although recent work suggests that if domains are sufficiently different, the performance
of the fine-tuned model will be limited \cite{yadlowsky_pretraining_2023}. The main point of comparison in this thesis will be the performance of
pretrained models fine-tuned on in-domain (aviation) data and models trained from scratch on in-domain data.

\subsection{Notation and Terminology}
This section is included for the sake of clarity throughout this thesis. In pursuit of that, a few terms and notations are defined below.

@@

\begin{itemize}
    \item A \textbf{word} is the representation in writing of a sound or combination of sounds that symbolizes and communicates a meaning
    \item A \textbf{sentence} is a grammatical unit that is syntactically independent that is expressed or understood and contains at least one finite
          verb
\end{itemize}

@@

\begin{itemize}
    \item A \textbf{token} shall be defined as the smallest, fundamental component of a sequence; these can be single characters, groups of
          characters, punctuation, etc. \cite{devlin_bert_2019,wu_googles_2016,sennrich_neural_2016}
    \item A \textbf{sequence} shall be defined as a set of tokens (distinct from a sentence)
\end{itemize}

\noindent
As mentioned above, we consider each sample in a corpus to be a sentence, then as the corpus is collated, processed, and tokenized, each sample will
be processed into sequences of tokens. These sequences are what will be analyzed and modeled by the language models.

@@

\subsection{Transformer Neural Networks}
% from thesis proposal; revise and add to this as more references are used
% transformer history (background)
The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved state-of-the-art
performance on language translation tasks \cite{vaswani_attention_2017}. Transformer architectures are extremely effective at learning representations
and understandings of languages to predict token probabilities instead of transforming one language into another
\cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been found to be very effective at other NLP tasks such as prompt completion
and sentiment analysis among others (i.e.~auto-regressive and sequence classification tasks, respectively)
\cite{lewis_bart_2019,radford_improving_2018}. Devlin et al.~\cite{devlin_bert_2019} proposed pretraining methods such as masked language modeling
and next sentence prediction to improve natural language understanding for downstream tasks, which was further expanded upon by Liu et
al.~\cite{liu_roberta_2019} by training longer, with larger batch sizes, and removing the next sentence prediction task.

@@

\section{Data Sources}\label{sec:data_source}
Four corpora are used in this thesis. The datasets are primarily speech corpora intended for automatic speech recognition research in aviation
communications, linguistics research in aviation English, or both. All data in all four corpora include both recordings and transcriptions, which are
audio and text, respectively. The transcriptions are used to pretrain and train the language models and tokenizers (described in sections
\ref{sec:language_models} and \ref{sec:tokenizers}).

@@

\section{Data Preparation}\label{sec:data_preparation}
All corpora listed above (see Section \ref{sec:data_source}) are created at different times, for various purposes, and by multiple authors. As a
result, the data in these corpora all have different formats. ATCC, for example, uses Lisp-style lists as below:

@@

\clearpage
\begin{lstlisting}
<data>
<segment>
<start>3.79</start>
<end>6.85</end>
<speaker>B</speaker>
<speaker_label>OK-PMB</speaker_label>
<text>level one hundred Oscar Kilo Papa Mike Bravo</text>
<tags>
<correct>0</correct>
<correct_transcript>1</correct_transcript>
<correct_tagging>0</correct_tagging>
<non_english>0</non_english>
</tags>
</segment>
</data>
\end{lstlisting}

\clearpage
ZCU CZ ATC uses a different nesting and labeling style, like below\footnote{Some information is excluded for readability.}:

@@

% TODO: fix; lost some text somehow
All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used,
    although these are built into most Python distributions by default.}. The text data is extracted from the corpus transcripts by reading each
transcript into the main memory of the system according to the format prescribed by the corpus. The text corresponding to the transcript is isolated,
copied, and stored to a separate region of memory (a Python list of strings) that corresponds to the original corpus. The resulting isolated
transcripts are written to files that correspond to the original corpus as well as combined, shuffled, and written to a file in the standard corpus
format (an ASCII text file with one line in the file corresponding to one sample from the corpus). The resulting output is four corpus files that
correspond to the original corpora in addition to a fifth corpus file with the samples from all four corpora combined and shuffled for later use.

@@

% 22 spelling errors
Manually reviewing every sample in the corpus is unrealistic, considering that there are over 50,000 data samples and would be subject to the same
human error that introduced those errors in the first place. Instead, the frequency of token occurrence in the corpora, combined with manual review
thereafter, was used to detect and correct the most common errors. The frequency of occurrence of a token is calculated using equation
(\ref{eq:token_frequency_occurrence}).

@@

\noindent
Where $f_w$ is the frequency at which a token occurs in a corpus, $\phi_w(w)$ is the number of times a token, $w$, occurs, and $D_{tokens}$ is the
total number of tokens in the corpus. Tokens with the lowest frequency across corpora (and lowest number of occurrences) were manually reviewed and
corrected where necessary. An $A \rightarrow B$ mapping system was used to correct the errors, where \(A\) is the erroneous token and \(B\) is the
corrected or replaced token. As the text is parsed from the corpora, each sample is analyzed. If an erroneous token, $A$, is found, it is replaced
with the corrected version, $B$. 22 unique errors were detected and subsequently corrected or removed using this method. However, multiple instances
of those errors typically occurred.

% hesitations, cut off words, different kinds of filled pauses
The mapping system described above is also used to modify corpora conventions to be consistent across corpora. Three areas are addressed here:
hesitations, cut-off, or partial words, and filled pauses within speech. These areas are defined differently between the four corpora, which makes
unifying the transcripts difficult. Hesitations, for example, are defined by one corpus as incomplete or non-English speech sounds and left completely
undefined by another (despite annotators making notes about hesitations). The way in which hesitations, pauses, and partial words are transcribed
differs between corpora as well. Two corpora prescribe special tokens to these phenomena, such as ``[hes]'', ``[hesitation]'', ``<pause>'', or
similar. In contrast, the others transcribe the data as it sounds (i.e., ``uh'', ``eh'', or ``er'' for filled pauses and ``cir-'' or ``cir+'' for
instances that were cut-off mid pronunciation). To facilitate the unification of the transcripts for all four corpora, the conventions for
transcribing these phenomena are redefined in Table \ref{tab:phenomena_definitions}.

\begin{table}[!t]
    \centering
    \begin{tabular}{l p{0.65\linewidth}}
        \toprule
        Phenomenon               & Definition                                                                                                              \\
        \midrule
        Hesitations              & Instances of hesitation for which the specific realization is not defined (usually some special token is used instead). \\
        \midrule
        Partial \& cut-off words & Partially pronounced words are indicated by a hyphen character where the pronunciation of the word is missing.          \\
        \midrule
        Filled pauses            & Hesitations with a verbal component that has been described or otherwise defined by the corpus/annotator.               \\
        \bottomrule
    \end{tabular}
    \caption{Speech/transcription phenomena with corresponding convention for transcribing or translating, depending on the corpus being processed.}
    \label{tab:phenomena_definitions}
\end{table}

@@

\subsection{Data Analysis}\label{sec:data_analysis}
% sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
Relevant corpora statistics were observed or calculated and are presented in Table \ref{tab:corpora_stats}. This includes the number of samples in the
corpus, the mean sequence length per sample, the number of unique tokens, and the total number of tokens in the corpus. The same statistics were also
calculated across corpora since all four corpora will be combined for training and testing the language and speech recognition models. Note that this
is based on the raw data after loading and unifying the corpora before preprocessing for model training/testing and outlier removal.

@@

@@

Sci-Kit Learn's \cite{pedregosa_scikit-learn_2011} implementation of the LOF algorithm was used with a value of \(K = 35\) for the K-Nearest Neighbors
algorithm and Minkowski distance for distance metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively used for distance
calculations). A total of 47 samples were detected as outliers and removed for analysis. They are summarized in Table \ref{tab:outlier_stats}.

@@

Although all of the outliers have longer sequence lengths than the typical sequence lengths of the other samples in the data, going through the
outlying samples manually and reviewing each reveals that, for the most part, the samples are typical albeit somewhat complex communications. Except
for one outlier, all of the samples labeled outlying by the LOF algorithm were kept in the data. Examples of the outliers are listed in Appendix A in
Section \ref{sec:appendix_a}, including the sample removed.

\section{Model Architectures}

\subsection{Automatic Speech Recognition}\label{sec:automatic_speech_recognition}
Automatic speech recognition (ASR) or simply speech recognition is the process of translating spoken language to text i.e.~transcribing spoken
language. There have been a myriad of approaches to ASR problems, including, but not limited to rule-based, machine learning, and deep learning
methods. For the purposes of this work, we will focus primarily on deep learning approaches to ASR tasks with a specific focus on transcribing
spoken language in aviation (following from the data accumulated, processed, and analyzed in sections \ref{sec:data_source},
\ref{sec:data_preparation}, and \ref{sec:data_analysis}).

\subsection{Input Representation}\label{sec:asr_input_representation}
Although the internal processing of data varies between architectures, the format of the models input and output is the same for the models described
below. ASR models have seen much success and performance gains for speech-related tasks such as speech and speaker recognition through the use of
Mel-frequency Cepstral Coefficients (MFCCs) \cite{sahidullah_design_2012}. Since ASR is not the main focus of this thesis, the process for computing
the MFCCs will only be described from a high level as it is described in \cite{sahidullah_design_2012}. Readers should refer to
\cite{sahidullah_design_2012} for a more detailed, matrix-based approach for computing MFCCs. Given an audio signal:

@@

\noindent
The MFCCs are then used as input to the ASR models. Each model outputs a probability distribution of characters or tokens (depending on the vocabulary
the ASR model is initialized with) over a segment of audio.

\subsection{Training Objective}\label{sec:asr_training_objective}
The ASR models listed below both use a Connectionist Temporal Classification (CTC) training objective \cite{graves_connectionist_2006}. For a
dataset, $D$, made up of inputs and labels, $x$ and $y$, respectively.

@@

\noindent
Where $N$ is the number of samples in the dataset, $D$. The goal of the ASR model training procedure is to predict the target, $y$, based on the input
sequence, $x$. This can also be written as maximizing the likelihood of $p(y|x)$. More formally, the CTC objective function can be written as

@@

The novelty of the CTC objective is in the fact that the audio signal does not need to be segmented in order to label and train a network. Instead,
the output layer of the network has $L + 1$ units, where $L$ is the alphabet/vocabulary of the network. At each time step in the audio signal, the
network will predict either a token from its vocabulary or a blank token, hence the extra unit in the output layer. Repeated predictions and blank
tokens are removed from the prediction, then scored against the ground truth label for the audio signal \cite{graves_connectionist_2006}.

\subsubsection{Jasper}\label{sec:jasper}
Jasper is a family of end-to-end convolutional neural networks (CNNs) developed by NVIDIA to replace traditional ASR models that use separately
learned components for each stage of the pipeline (acoustic, pronunciation, language modeling, etc.) \cite{li_jasper_2019}.

Jasper models are designed with $B$ blocks and $R$ sub-blocks and are named accordingly as \textit{Jasper $B\mbox{x}R$} models. The Jasper
architecture also contains four extra convolutional blocks, one for preprocessing and three for postprocessing (see figure \ref{fig:jasper}). Each
convolutional sub-block contains one dimensional convolutions. Li et al.~\cite{li_jasper_2019} trained and tested a Jasper 10x5 model that produced
state-of-the-art results on the LibriSpeech, Wall Street Journal, and Fisher+Switchboard speech corpora.

@@

\subsubsection{QuartzNet}\label{sec:quartznet}
The QuartzNet architecture is based on Jasper with some key modifications, namely, that the one dimensional convolutions are replaced with ``one
dimensional time-channel separable convolutions''. As with Jasper, QuartzNet models are denoted as \textit{QuartzNet $B\mbox{x}R$}, however, blocks
are repeated $S$ times and have a certain number of input and output channels, $c_{in}$ and $c_{out}$, respectively.

@@

\subsection{Language Models}\label{sec:language_models}
% bert, roberta
% n-gram
This section introduces the chosen language model architectures and their pretrained checkpoints. Two transformer-based architectures, BERT and
RoBERTa, and one statistical/machine learning-based approach: N-gram language modeling.

>>>>

\begin{document}
% Riddle thesis title page
\begin{center}
    \LARGE
    \textbf{
        \textit{
            Spoken Language Processing and Modeling for Aviation Communications\\
        }
    }

@@

\section{Abstract}\label{sec:abstract}
With recent advances in machine learning and deep learning technologies and the creation of larger aviation-specific corpora, applying
natural language processing technologies, especially those based on transformer neural networks, to aviation communications is becoming
increasingly feasible. Previous work has focused on machine learning applications to natural language processing, such as N-grams and
word lattices. This thesis experiments with a process for pretraining transformer-based language models on aviation English corpora and
compare the effectiveness and performance of language models transfer learned from pretrained checkpoints and those trained from their
base weight initializations (trained from scratch). The results suggest that transformer language models trained from scratch outperform
models fine-tuned from pretrained checkpoints. The work concludes by recommending future work to improve pretraining performance and
suggestions for downstream, in-domain tasks such as semantic extraction, named entity recognition (callsign identification), speaker
role identification, and speech recognition.

\section{Introduction}\label{sec:introduction}
There have been noticeable and significant advancements in the general domain of artificial intelligence (AI) recently. Natural
language processing and language modeling technologies, in particular, have made tremendous advances over the past two decades
\cite{hannun_deep_2014,lee_unsupervised_2009,li_jasper_2019,kriman_quartznet_2020}, achieving performance levels good enough to be brought to
consumer-facing products such as personal assistants, search engines, phones, televisions, etc. Due to the success in the general domain, applications
have begun to spin off into problem-specific domains, such as (and most notably for this work) aviation. For example, several speech recognition
systems with rule- and machine learning-based natural language processing, inference, and semantic extraction algorithms have been implemented at
EUROCONTROL simulation centers in Europe to train air traffic controllers and study the potential effects of AI-augmented air traffic management (ATM)
systems on controller workloads \cite{helmke_quantifying_2017,helmke_assistant-based_2015}. Preliminary studies have found that AI-augmented ATM
systems reduce controller workloads and fuel consumption of aircraft by optimizing aircraft routes in the airspace and reducing departure/arrival
times, thereby reducing fuel consumption \cite{helmke_quantifying_2017}.

At the time of this writing, the current focus of general machine learning and deep learning algorithms in the language modeling domains has been on a
variety of deep learning methods/architectures, namely, recurrent, convolutional, and transformer neural networks, due to the availability of large
natural language corpora such as WikiText \cite{merity_pointer_2016} and IMDB \cite{maas_learning_2011}. In contrast to the general domain, the
aviation domain still prefers and sees much success with traditional machine learning models for language modeling applications. Typically, the most
prevalent of the traditional models are N-grams, word lattices, and rule-based approaches, depending on the application and operating environment of
the models. The most likely reason for this divergence between the domains is the difference and relative lack of labeled data in the aviation domain.
By comparison, the largest corpora in the general domain are either WikiText or the BookCorpus, which contain trillions of tokens, whereas, for
aviation, the largest corpus is Air Traffic Control Complete with approximately 26 hours of labeled data amounting to just over 300,000
tokens\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are labeled.}. This disparity
in data availability is exacerbated by the fact that aviation English uses a specialized vocabulary in addition to specialized pronunciation for some
words to enhance understandability in a radiotelephonic (R/T) medium \cite{paltridge_handbook_2013}. This makes the data labeling process
significantly more difficult because it not only requires two types of data (audio and text) to create labels, but also requires labelers to have
domain knowledge and experience to create effective transcriptions. This, in addition to the following facets, further contributes to the difficulty
of transcription: (1) the medium of communication, (2) the environment in which the communication occurs\footnote{For example, a noisy and possibly
    low fidelity medium, depending on the equipment used to transmit and receive communications}, (3) a noisy and active communication environment
i.e.~an active flight deck under the influence of wind, engine noise, etc., (4) an active Air Traffic Control (ATC) tower and channels contributing to
increased background noise on the ATC side, and (5) a rapid rate of speech. Pilots are also operating sophisticated equipment during while
communicating, requiring significant attention and mental resources further contributing to the complexity of the communications
\cite{paltridge_handbook_2013}.

@@

The key component missing for both of these applications is an effectively pretrained language model that can be used for these downstream tasks, so
the goal of this thesis is to run several experiments using transformer-based language models to determine the most effective strategy for pretraining
the model for downstream tasks. Recommended best practice is to use a pre-existing language model checkpoint and fine-tune on additional data
\cite{devlin_bert_2019,liu_roberta_2019,lewis_bart_2019}, although recent work suggests that if domains are sufficiently different, the performance
of the fine-tuned model will be limited \cite{yadlowsky_pretraining_2023}. The main point of comparison in this thesis will be the performance of
pretrained models fine-tuned on in-domain (aviation) data and models trained from scratch on in-domain data. The experiments and results presented in
this thesis are preliminary applications of end-to-end deep learning solutions to ASR and NLP problems in aviation. The intention is for this work to
provide knowledge and possibly checkpoints for the ASR and language models which can be used for downstream tasks such as callsign detection and beam
search prediction decoding for ASR models.

\subsection{Notation \& Terminology}
This section is included for the sake of clarity throughout this thesis. In pursuit of that, a few terms and notations are defined below.

@@

\begin{itemize}
    \item A \textbf{word} is the representation in writing of a sound or combination of sounds that symbolizes and communicates a meaning.
    \item A \textbf{sentence} is a grammatical unit that is syntactically independent that is expressed or understood and contains at least one finite
          verb.
\end{itemize}

@@

\begin{itemize}
    \item A \textbf{token} shall be defined as the smallest, fundamental component of a sequence; these can be single characters, groups of
          characters, punctuation, etc. \cite{devlin_bert_2019,wu_googles_2016,sennrich_neural_2016}
    \item A \textbf{sequence} shall be defined as a set of tokens (distinct from a sentence).
\end{itemize}

\noindent
As mentioned above, we consider each \textbf{sample} in a corpus to be a sentence, then as the corpus is collated, processed, and tokenized, each
sample will be processed into sequences of tokens. These sequences are what will be analyzed and modeled by the language models.

@@

\subsection{Transformer Neural Networks}
% from thesis proposal; revise and add to this as more references are used
% transformer history (background)
The transformer neural network architecture (shown in Figure \ref{fig:transformer_arch}) was proposed in 2017 for Neural Machine Translation tasks and
immediately achieved state-of-the-art performance on language translation tasks \cite{vaswani_attention_2017}. Transformer architectures are extremely
effective at learning representations and understandings of languages to predict token probabilities instead of transforming one language into another
\cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been found to be very effective at other NLP tasks such as prompt completion
and sentiment analysis among others (i.e.~auto-regressive and sequence classification tasks, respectively)
\cite{lewis_bart_2019,radford_improving_2018}. Devlin et al.~\cite{devlin_bert_2019} proposed pretraining methods such as masked language modeling
and next sentence prediction to improve natural language understanding for downstream tasks, which was further expanded upon by Liu et
al.~\cite{liu_roberta_2019} by training longer, with larger batch sizes, and removing the next sentence prediction task.

\begin{figure}[!t]
    \centering
    \includegraphics[width=0.5\linewidth]{figures/transformer_arch.png}
    \caption{The transformer neural network architecture. Copied from Vaswani et al.~\cite{vaswani_attention_2017}}
    \label{fig:transformer_arch}
\end{figure}

@@

\section{Datasets \& Preparation}\label{sec:data_source}
Four corpora are combined and used for the experiments in this thesis. The datasets are primarily speech corpora intended for automatic speech
recognition research in aviation communications, linguistics research in aviation English, or both. All data in all four corpora include both
recordings and transcriptions, which are audio and text, respectively. The transcriptions are used to pretrain and train the language models and
tokenizers (described in sections \ref{sec:language_models} and \ref{sec:tokenizers}).

@@

\subsection{Data Preparation}\label{sec:data_preparation}
All corpora listed above (see Section \ref{sec:data_source}) are created at different times, for various purposes, and by multiple authors. As a
result, the data in these corpora all have different formats. ATCC, for example, uses Lisp-style lists as below:

@@

\begin{lstlisting}
<data>
<segment>
<start>3.79</start>
<end>6.85</end>
<speaker>B</speaker>
<speaker_label>OK-PMB</speaker_label>
<text>level one hundred Oscar Kilo Papa Mike Bravo</text>
<tags>
<correct>0</correct>
<correct_transcript>1</correct_transcript>
<correct_tagging>0</correct_tagging>
<non_english>0</non_english>
</tags>
</segment>
</data>
\end{lstlisting}

ZCU CZ ATC uses a different nesting and labeling style, like below\footnote{Some information is excluded for readability.}:

@@

All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used,
    although these are built into most Python distributions by default.}. The text data is extracted from the corpus transcripts by reading each
transcript into the main memory of the system according to the format prescribed by the corpus. The text corresponding to the transcript is isolated,
copied, and stored to a separate region of memory (a Python list of strings) that corresponds to the original corpus. The resulting isolated
transcripts are written to files that correspond to the original corpus as well as combined, shuffled, and written to a file in the standard corpus
format (an ASCII text file with one line in the file corresponding to one sample from the corpus). The resulting output is four corpus files that
correspond to the original corpora in addition to a fifth corpus file with the samples from all four corpora combined and shuffled for later
use\footnote{See the \lstinline|parse_transcripts| functions in the files at \url{https://github.com/AVanDeBrook/msece-thesis/tree/main/source/data}
    for implementation details.}.

@@

% 22 spelling errors
Manually reviewing every sample in the corpus is unrealistic, considering that there are over 50,000 data samples and would be subject to the same
human error that introduced those errors in the first place. Instead, the frequency of token occurrence in the corpora, combined with manual review
thereafter, was used to detect and correct the most common errors. Tokens with the lowest frequency across corpora (and lowest number of occurrences)
were manually reviewed and corrected where necessary. Tokens that needed to be corrected typically occurred five times or less per corpus although
all tokens with occurrences ranging from 1-50 were reviewed. The relative frequency of those tokens varies by corpus (due to the differing amont of
tokens in total for each corpus) The frequency of occurrence of a token is calculated using equation (\ref{eq:token_frequency_occurrence}), below

@@

\noindent
where $f_w$ is the frequency at which a token occurs in a corpus, $\phi_w(w)$ is the number of times a token, $w$, occurs, and $D_{tokens}$ is the
total number of tokens in the corpus.

An $A \rightarrow B$ mapping system was used to correct the errors, where \(A\) is the erroneous token and \(B\) is the
corrected or replaced token. As the text is parsed from the corpora, each sample is analyzed. If an erroneous token, $A$, is found, it is replaced
with the corrected version, $B$. 22 unique errors were detected and subsequently corrected or removed using this method. However, multiple instances
of those errors typically occurred.

% hesitations, cut off words, different kinds of filled pauses
The mapping system described above is also used to make corpora conventions to be consistent across corpora. Three areas are addressed here:
hesitations, cut-off, or partial words, and filled pauses within speech. These areas are defined differently between the four corpora, which makes
unifying the transcripts difficult. Hesitations, for example, are defined by one corpus as incomplete or non-English speech sounds and left completely
undefined by another (despite annotators making notes about hesitations). The way in which hesitations, pauses, and partial words are transcribed
differs between corpora as well. Two corpora prescribe special tokens to these phenomena, such as ``[hes]'', ``[hesitation]'', ``<pause>'', or
similar. In contrast, the others transcribe the data as it sounds (i.e., ``uh'', ``eh'', or ``er'' for filled pauses and ``cir-'' or ``cir+'' for
instances that were cut-off mid pronunciation). To facilitate the unification of the transcripts for all four corpora, the conventions for
transcribing these phenomena are redefined in Table \ref{tab:phenomena_definitions}.

\begin{table}[!t]
    \centering
    \begin{tabular}{l p{0.65\linewidth}}
        \toprule
        Phenomenon               & Definition                                                                                                                \\
        \midrule
        Hesitations              & Instances of spoken words for which the specific realization is not defined (usually some special token is used instead). \\
        \midrule
        Partial \& cut-off words & Partially pronounced words are indicated by a hyphen character where the pronunciation of the word is missing.            \\
        \midrule
        Filled pauses            & Hesitations with a verbal component that has been described or otherwise defined by the corpus/annotator.                 \\
        \bottomrule
    \end{tabular}
    \caption{Speech/transcription phenomena with corresponding convention for transcribing or translating, depending on the corpus being processed.}
    \label{tab:phenomena_definitions}
\end{table}

@@

\subsection{Data Analysis}\label{sec:data_analysis}
% sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
To better understand the above four corpora, we calculate and tabulate their relevant corpora statistics in Table \ref{tab:corpora_stats}. This
includes the number of samples in the corpus, the mean sequence length per sample, the number of unique tokens, and the total number of tokens in the
corpus. The same statistics were also calculated across corpora since all four corpora will be combined for training and testing the language and
speech recognition models. Note that this is based on the raw data after loading and unifying the corpora before preprocessing for model
training/testing and outlier removal.

@@

\noindent
In the equation above, $V_1$ and $V_2$ are the two numbers being compared.

@@

Sci-Kit Learn's \cite{pedregosa_scikit-learn_2011} implementation of the LOF algorithm was used with a value of $K = 35$ for the K-Nearest Neighbors
algorithm; the value for $K$ is arbitrarily chosen, partially through trial and error and the size of the clusters for sequence lengths. Minkowski
distance is used for the distance metrics and is defined as

\begin{equation}\label{eq:minkowski_distance}
    d(x, y) = (|x - y|^p)^{\frac{1}{p}}
\end{equation}

\noindent
where $x$ and $y$ are the points between which the distance is being measured and $p$ is the order on which to calculate the distance\footnote{For
    the LOF algorithm, $p$ can be arbitrarily chosen.}. $p=2$ is chosen in this case to make the Minkowski distance equivalent to Euclidean distance:

\begin{equation}\label{eq:euclidean_distance}
    d(x, y) = \sqrt{|x-y|^2}
\end{equation}

\noindent
A total of 47 samples were detected as outliers and removed for analysis. They are summarized in Table \ref{tab:outlier_stats}.

@@

Although all of the outliers have longer sequence lengths than the typical sequence lengths of the other samples in the data, going through the
outlying samples manually and reviewing each reveals that, for the most part, the samples are typical albeit somewhat complex communications. Except
for one outlier, all of the samples labeled outlying by the LOF algorithm were kept in the data. Examples of the outliers are listed in Appendix A in
Section \ref{sec:appendix_a}, including the removed sample.

\section{End-to-End ASR Models \& Experiment Results}
Automatic speech recognition (ASR) or simply speech recognition is the process of translating spoken language to text i.e.~transcribing spoken
language. There have been a myriad of approaches to ASR problems, including, but not limited to rule-based, machine learning, and deep learning
methods. For the purposes of this work, we will focus primarily on end-to-end deep learning approaches to ASR tasks with a specific focus on
transcribing spoken language in aviation (following from the data accumulated, processed, and analyzed in sections \ref{sec:data_source},
\ref{sec:data_preparation}, and \ref{sec:data_analysis}).

\subsection{Input Representation}\label{sec:asr_input_representation}
Although the internal processing of data varies between architectures, the representation of the models' input for the models described below.
ASR models have seen much success and performance gains for speech-related tasks such as speech and speaker recognition through the use of
Mel-frequency Cepstral Coefficients (MFCCs) \cite{sahidullah_design_2012}. Since ASR is not the main focus of this thesis, the process for computing
the MFCCs will only be described from a high level as it is described in \cite{sahidullah_design_2012}. Readers should refer to
\cite{sahidullah_design_2012} for a more detailed, matrix-based approach for computing MFCCs. Given an audio signal:

@@

\noindent
The MFCCs are then used as input to the ASR models. It should be noted that MFCCs are not the only input representation available
for ASR models, just the most popular choice among the models used here. Each model outputs a probability distribution of characters or tokens
(depending on the vocabulary the ASR model is initialized with) over a segment of audio.

\subsection{Training Objective}\label{sec:asr_training_objective}
The ASR models listed below both use a Connectionist Temporal Classification (CTC) training objective \cite{graves_connectionist_2006}. For a
dataset, $D$, made up of samples and labels, $x$ and $y$, respectively, defined as

@@

\noindent
where $N$ is the number of samples in the dataset. The goal of the ASR model training procedure is to predict the target, $y$, based on the input
sequence, $x$. This can also be written as maximizing the likelihood of $p(y|x)$. More formally, the CTC objective function can be written as

@@

The novelty of the CTC objective is in the fact that the MFCC frames do not need to be aligned with the letter-based tokens to train a network.
Instead, the output layer of the network has $L + 1$ units, where $L$ is the alphabet/vocabulary of the network. At each time step in the audio
signal, the network will predict either a token from its vocabulary or a blank token, hence the extra unit in the output layer. Repeated predictions
and blank tokens are removed from the prediction, then scored against the ground truth label for the audio signal \cite{graves_connectionist_2006}.
Note that Graves et al.~\cite{graves_connectionist_2006} defined the tokens predicted at each time step as letters and the vocabulary,
correspondingly, as the letters \textit{A} through \textit{Z} of the alphabet.

\subsection{Jasper}\label{sec:jasper}
Jasper is a family of deep convolutional neural networks (CNNs) developed by NVIDIA as end-to-end ASR models to replace traditional ASR models that
use separately learned components for each stage of the pipeline (acoustic, pronunciation, language modeling, etc.) \cite{li_jasper_2019}.

Jasper models, shown in Figure \ref{fig:jasper} are designed with $B$ blocks and $R$ sub-blocks and are named accordingly as
\textit{Jasper $B\mbox{x}R$} models. The Jasper architecture also contains four extra convolutional blocks, one for preprocessing, shown at the
bottom of Figure \ref{fig:jasper}, and three for postprocessing, shown at the top of Figure \ref{fig:jasper}. Each convolutional sub-block contains
one dimensional convolutions. Li et al.~\cite{li_jasper_2019} trained and tested a Jasper 10x5 model that produced state-of-the-art results on the
LibriSpeech, Wall Street Journal, and Fisher+Switchboard speech corpora.

@@

\subsection{QuartzNet}\label{sec:quartznet}
The QuartzNet architecture is based on Jasper with some key modifications, namely, that the one dimensional convolutions are replaced with ``one
dimensional time-channel separable convolutions'', as shown in Figure \ref{fig:quartznet}. As with Jasper, QuartzNet models are denoted as
\textit{QuartzNet $B\mbox{x}R$} with $B$ and $R$ representing the number of blocks and sub-blocks, respectively. However, blocks are repeated $S$
times and have a certain number of input and output channels, $c_{in}$ and $c_{out}$, respectively. Time-channel separable convolutions are separated
into two components: (1) convolutional layers that operate on each audio channel independently across time frames and (2) convolutional layers that
operate on time frames independently across all audio channels \cite{kriman_quartznet_2020}.

@@

\subsection{Experiment Results}
This section will summarize the experiment setup, evaluation metrics, and final results of the two best performing ASR models. Two Jasper models with 10
blocks and 5 sub-blocks and two QuartzNet models with 15 blocks and 5 sub-blocks were trained from scratch and fine-tuned from pretrained
checkpoints; four models were evaluated in total. Each model used a vocabulary of 28 tokens; the English alphabet, an apostrophe, and a blank label.
